---
title: "Chi squared test"
author: "Sathya"
date: "2024-12-04"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## LOading the dataset

```{r}
data <- read.csv("C:/Users/DELL/Downloads/archive (21)/PCOS_extended_dataset.csv")
head(data)
```

## Describing the variables 
PCOS..Y.N.is used to find the presence of PCOS zero means the absence and one means presence of PCOS.
Hair. loss.Y.N.is used to find the presence of hairloss zero means the absence and one means presence of hairloss.
##Summarizing the variables

```{r}
# Summarize levels and frequencies for PCOS
pcos_summary <- table(data$`PCOS..Y.N.`)
print(paste("PCOS (Y/N) Levels and Frequencies:"))
for (level in names(pcos_summary)) {
  print(paste(level, ":", pcos_summary[level]))
}

# Summarize levels and frequencies for Hairloss
hairloss_summary <- table(data$`Hair.loss.Y.N.`)
print(paste("Hairloss (Y/N) Levels and Frequencies:"))
for (level in names(hairloss_summary)) {
  print(paste(level, ":", hairloss_summary[level]))
}
```
##Printing frequency tables
```{r}
# Frequency tables
pcos_freq <- table(data$`PCOS..Y.N.`)
hairloss_freq <- table(data$`Hair.loss.Y.N.`)

# Print frequencies
print("PCOS Frequency Table:")
print(pcos_freq)


print("Hairloss Frequency Table:")
print(hairloss_freq)

```
##Visualizations using bar plots 
```{r}
barplot(pcos_freq, 
        main = "Distribution of PCOS", 
        xlab = "PCOS (Y/N)", 
        ylab = "Count", 
        col = "orange", 
        ylim = c(0, max(pcos_freq) + 10))
```

```{r}
barplot(hairloss_freq, 
        main = "Distribution of Hairloss", 
        xlab = "Hairfall (Y/N)", 
        ylab = "Count", 
        col = "skyblue", 
        ylim = c(0, max(hairloss_freq) + 10))
```
##Hypothesis:
Null Hypothesis (H₀): There is no relationship between PCOS and the hairloss.PCOS doesn't cause hairloss (the two variables are independent).
Alternative Hypothesis (H₁): There is a relationship between PCOS and the hairloss.PCOS can likely cause the hairloss (the two variables are not independent).

```{r}
# Create a contingency table
contingency_table <- table(data$`PCOS..Y.N.`, data$`Hair.loss.Y.N.`)
print("Contingency Table:")
print(contingency_table)
```

```{r}
# Perform Chi-Square Test of Independence
chi_square_result <- chisq.test(contingency_table)

# Output the test result
print("Chi-Square Test Result:")
print(chi_square_result)
```
so the x-square = 65.726 , df =1 and p-value is 5.183e-16

Conclusion:Since the p-value (5.183×10 −16 ) is significantly less than 0.05, reject the null hypothesis (H₀).

Implications:
So,There is very strong statistical evidence to suggest a relationship between PCOS and hairloss.
This means that having PCOS is significantly associated with the hairloss.

Reflections:
The Chi-Square test revealed a significant relationship between PCOS (Polycystic Ovary Syndrome) and Hairloss in the dataset, as evidenced by a very low p-value (p<0.05) and a high Chi-Square statistic (65.726). This indicates that the occurrence of hair loss is not independent of the presence of PCOS, suggesting a meaningful association between these two variables.The contingency table likely showed an uneven distribution of hair loss between individuals with and without PCOS. This uneven distribution reinforces the observed relationship, with a higher proportion of individuals with PCOS reporting hair loss.
If the majority of the individuals with PCOS have reported hair loss, it validates the connection between these two factors.
